Recruiting Chapter

### 2021-06-26 text/manuscript_june_2021.Rmd
copy-edit draft

---
title: "Recruiting Chapter"
author: 
  - Ozan Jaquette
  - Crystal Han
  - Irma Castaneda
bibliography: ./../other/burd_recruiting_chapter.bib
output: 
  bookdown::pdf_document2:
    toc: FALSE
csl: ./../other/apa.csl
urlcolor: blue
header-includes:
      - \usepackage{pdflscape}
      - \usepackage{geometry}
      - \usepackage{floatrow}
      - \usepackage{setspace}
      #- \doublespacing
      - \floatsetup{capposition=top}
fontsize: 11pt
---

@@

Prior to the pandemic, admissions counselors from colleges and universities traveled to local high schools each year in order to meet prospective students and maintain relationships with guidance counselors at feeder schools [@RN3519]. @RN4721 analyzed off-campus recruitiving visits made in 2017 by 15 public research universities. Surprisingly, 12 of the 15 universities made more visits to *out-of-state* high schools than *in-state* high schools. These out-of-state visits focused on public schools in affluent, predominantly white communities. Additionally, most public research universities in our sample made a disproportionate number of visits to out-of-state private high schools. For the University of California-Berkeley, for example, 155 (43%) of the 360 out-of-state high school visits were to private high schools. For the University of Colorado-Boulder, 321 (36%) of the 901 out-of-state high school visits were to private high schools.

@@

The mission of selective private colleges and universities is more squarely associated with social reproduction rather than social mobility. Historically, these institutions depend on patronage from philanthropic interests -- in connection with business and religious interests -- and depend from wealthy households who provide tuition revenue and donations [@RN4517]. In turn, selective private institutions contribute to social reproduction by providing access to their patrons. The "Chetty" data on parental income by college show that the overwhelming majority of students at selective private institutions have parental income in the top quintile [@RN4722; @RN4723]. A significant caveat is that religiously affiliated universities may prioritize enrollment opportunities for students of the same faith regardless of wealth, but even here the mission is to benefit families from a particular cultural status group [CITE].

The traditional, albeit mythologized, public good mission of public research universities is tied to social mobility and development of the state [@RN2269; @RN921]. Public research universities are designated the unique responsibility of educating the future civic, professional, and business leaders of the state. Quoting 19th century University of Michigan President James Angell, these institutions contribute to social mobility by providing “an uncommon education for the common man” [as cited in @RN3608, p. 279] who could not afford tuition at elite private institutions.

We argue that the recruiting behavior of colleges and universities is an indicator of organizational enrollment priorities and, in turn, the extent to which enrollment priorities serve social mobility or social reproduction. Recruiting visits to private high schools are consistent with the social reproduction mission of selective private universities, regardless of whether the university is targeting a particular high school because of its academic reputation, religious affiliation, or the affluence of its student body. By contrast, for public research universities, visits to private high schools -- particularly out-of-state schools -- are antithetical to mission of social mobility for high-achieving, low-income state residents.

Analyses of recruiting visits to private high schools can contribute novel insights about the "privatization" of public research universities. Within the scholarly literature on privatization, one strand of scholarship conceptualizes privatization as a process whereby declines in government funding cause public institutions to grow alternative revenue sources, particularly revenue from tuition and research commercialization [@RN4543@; @RN2101]. Of relevance to this chapter, public research universities responded to declines in state appropriations by growing enrollment from nonresident students [@RN3753], who typically pay two- to three-times more than resident students. A second strand within the privatization literature conceptualizes privatization as a process whereby public institutions become more similar -- in terms of characteristics or behavior -- to for-profit or private non-profit organizations [@RN4543]. A weakness of this literature is that few empirical studies analyze public institutions on a behavior associated with private institutions.

Therefore, this chapter compares public research universities to selective private universities on the dimension of recruiting visits to private high schools, a behavior associated with the social reproduction orientation of selective private universities. Our primary goal is to learn about public research universities, using the behavior of private universities as a benchmark to develop more thoughtful insights about public research universities.  We analyze recruiting visits made in 2017 calendar year by a sample of 15 public research universities, 14 selective private universities, and 12 selective private liberal arts colleges.

Because little is known about recruiting at private high schools by public universities, we conduct descriptive analyses that seek to identify basic patterns and relationships. Because a recruiting visit to a private high school is an indicator of a relationship between two actors, we also utilize social network methods in order to compare the recruiting networks of public universities to those of private universities. Analyses are informed by the following three research questions:

@@

SHORT PARAGRAPH GIVING OUTLINE OF CHAPTER

@@

## Situating Off-Campus Recruiting Within Enrollment Management

@@

Institutions identify undergraduate "prospects" by purchasing "student lists" -- containing contact, demographic, and academic achievement information -- from College Board, ACT, and other vendors. Universities control which prospects are contained within a list by filtering on criteria such as zip code, test scores, and high school GPA. Once prospects and inquiries are identified, they are targeted with remote and face-to-face recruiting interventions designed to solicit applications and deepen engagement. Remote recruiting interventions include postcards, brochures, emails, text messages, and targeted social media. Face-to-face recruiting interventions include on-campus visits by prospects and off-campus recruiting visits by admissions representatives to high schools, community colleges, college fairs, etc. Institutions utilize advertising (traditional and digital) and social media to solicit inquiries and to create positive "buzz" amongst prospects [@RN4411]<!--[CITE OTHER]-->. Given the the rise in "stealth applicants" who do not inquire before applying (Dupaul & Harris, 2012), advertising and social media enables universities to tell their story to prospects who do not wish to be contacted.

@@

## Empirical Scholarship from Sociology

@@

Empirical academic scholarship on off-campus recruiting is mostly limited to a modest number of sociological case studies, which typically analyze off-campus recruiting as part of a broader analysis of enrollment management [e.g., @RN4324; @RN3519; @RN4538; @RN4520; @RN4407]. @RN4324 analyzes visits from the perspective of students at two racially and socioeconomically diverse public high schools. @RN4324 found that high school visits -- including college fairs, instant decision events, and small-group representative visits -- influenced where students applied and where they enrolled. This finding was strongest for first-generation students and under-represented students of color, who often reported that "school counselors had low expectations for them and were too quick to suggest that they attend community college" (p. XXX) This trust vacuum created an opportunity for colleges because these students were drawn to colleges that connected with them and made them feel wanted.  For example, @RN4324 describes a high-achieving, first-generation, African American student who was admitted by a highly selective liberal arts college but chose to attend a less selective college that "seemed to want him more" (p. XXX). By contrast, affluent students with college educated parents were less taken by such overtures and more concerned with college prestige.

@RN3519, an ethnography of the admissions office at a selective private liberal arts college, highlights the relational function of off-campus recruiting visits. Stevens [-@RN3519, p. 54] states that "the College's reputation and the quality of its applicant pool are dependent upon its connections with high schools nationwide." Therefore, during the autumn "travel season," admissions officers visit selected high schools across the country "to spread word of the institution and maintain relationships with guidance counselors" (p. 53-54). The College tended to visit the same "feeder" schools year after year because recruiting depends on long-term relationships with high schools. The high schools they visited tend to be affluent schools -- in particular, private schools -- that enroll high-achieving students who can afford tuition and had the resources and motivation to host a successful visit. Whereas @RN4402 highlights the effect of recruiting visits on inquiries and enrollees, findings from @RN3519 suggest that the College may have valued recruiting visits primarily as a means of maintaining relationships with guidance counselors. From this perspective, recruiting visits may affect outcomes such as inquiries, applications, and matriculation through their affect on high school guidance counselors. The logic is that a guidance counselor who views a college favorably will steer students to the college.

@@

The horsetrading described by @RN4407 depends on a relationship where the college can trust statements made by the high school counselor and vice-versa. This relationship is the product of repeated interactions over many years. A high school that makes false statements about applicant intentions faces consequences. The college "might stop taking these telephone calls and ignore the information provided. They may even start accepting fewer students from the school, thinking it is less than an honest" [@RN4321, p. 106]. Because the college and the high school are mutually dependent, however, both "have an incentive to continue with a strong, honest relationship" [@RN4321, p. 106]. Although, such horsetrading may be less pervasive now than it was in prior decades and may be less common at non-elite private schools. Additionally, college admissions officers from public universities likely have less authority to engage in these tacit negotiations because admissions criteria at public universities face greater public scrutiny than admissions criteria at private colleges and universities.

@@

@RN3519 and Khan [-@RN4321;-@RN4407] suggest that strong relationships are mutually important for the college and for the private high school. Strong relationships enable colleges and schools to negotiate and send trustworthy information to one another. Without a strong relationship, it unlikely that a college admissions counselor will "take the call" of a high school guidance counselor [@RN4321; @RN4407]. Even in the absence of horsetrading described by @RN4407, relationships enable the college admissions counselor and the high school guidance counselor to tell one another "their story" and relay that story to constituents [@RN3519]. The college admissions counselor explains why the college/university is a "special place" that high school students should want to attend [@RN3519]. The guidance counselor explains why the high school is a special place, that even students with lower grades have outstanding extracurricular strengths that will benefit the college [@RN4321].

@@

Having conceptualized a recruiting visit as an indicator of a social relation, this section introduces basic concepts from social network analysis and motivate the research questions that guide empirical analyses.

@@

Whereas "one-mode" networks consist of vertices of the same "type" (e.g., in a publication network each vertex is an author), "two-mode" networks consist of vertices associated with one type of actor/entity having connections to vertices of another type. For example, an actor-movie network consists of actors (mode 1) who appear in movies (mode 2), and an actor shares an edge with a movie if the actor appears in the movie. The most commonly analyzed two-mode network in social networks literature is a corporate board-director network  [e.g., @RN4663]. These networks consists of directors (mode 1) and organizational boards (mode 2) and a particular director shares an edge with a particular organizational board if they sit on that board. Similarly, our recruiting visit data form a two-mode network, consisting of private high schools (mode 1) and college (mode 2). An off-campus recruiting visit from a college/university to a private high school is the "edge" that connects two particular vertices. In this network, edges can only occur between vertices of a different type; that is a college can visit a high school, but colleges cannot visit other colleges and high schools cannot receive visits from other high schools.

@@

In social network analysis, *centrality* refers to the importance or prominence or "being in the middle of things" of an actor/vertex in a social network [CITE]. Many alternative measures of centrality have been developed. All measures agree that centrality is an actor/vertex-level construct (e.g., each college in our network will have a centrality value, regardless of which centrality measure we use). However, different measures of centrality are based on different theoretical conceptualizations -- for example, being directly connected to the most actors versus being the intermediary between otherwise disconnected actors -- and are operationalized using different mathematical algorithms.

The simplest measure of centrality, *degree centrality* measures the number of edges directly connected to a vertex. For each college in our network, degree centrality is simply the number of different private high schools visited by the college. For each high school in our network, degree centrality is simply the number of colleges in our network that visited the high school. Social network theory conceptualizes degree centrality as a measure of "local" centrality, capturing the number of direct connections for each actor/vertex. For our empirical context, we conceive of degree centrality as a measure of scale of the number of private high schools a college visits and, equivalently, the number of private high schools we observe haging a social relation with each college. Given that we conceptualize visits to private high schools as a behavior consistent with the social reproduction orientation of selective private colleges/universities, we are interested in comparing how scale differs between public and private colleges/universities. This interest motivates research question 1, "how does the scale of visits to private high schools by public research universities compare to the scale of visits to private high schools by selective private universities?" We answer this research question by comparing degree centrality across universities.

@@

Degree is a limited measure of centrality in that only identifies direct connections. By contrast, *k-path centrality* centrality measures the extent to which an actor/vertex is indirectly connected to others, *betweenness centrality* measures the extent to which a vertex is an intermediary between otherwise disconnected groups of actors, and *eigenvector centrality* measures the extent to which a vertex is connected to highly connected others. Betweenness centrality tends to be relevant for empirical contexts (e.g., which manager gets credit for an idea) where an actor who occupies a position between groups can benefit by being a broker of information/ideas [@RN975]. By contrast, this chapter is more concerned with whether a particular college tends to have indirect connections with some colleges but not others, and what is the subset of high schools this group of colleges tends to visit. Research question 3, motivated below, explores these ideas.

@@

A common finding in social networks research is that actors are more likely to form connections with actors who are are similar to them [@RN1722]. Said differently, ego networks are biased towards "homophily" rather than "heterophily." Homophily is the idea that two vertices are more likely to be connected if they have similar characteristics (e.g., two people are more likely to be friends on Facebook if they have similar political ideologies). Heterophily, the opposite of homophily, is when pairs of connected vertices have different characteristics from one another.

In our analyses of recruiting visits from colleges to private high schools, we are interested in the existence of homophily versus heterophily for two reasons. First, prior scholarship suggests that private colleges and private high schools are more likely to have a relationship when they share common ideologies or view themselves as belonging to the same echelon [@RN3519; @RN4321;@RN4407]. Second, with respect to our substantive interest the behavior of public research universities, the characteristics of visited private high schools have consequences for subsequent enrollment composition of public research universities. Therefore, we are interested in whether universities tend to visit high schools that are similar to them on certain characteristics. Additionally, we are interested in whether public research universities favor high schools with certain characteristics, independent of homophily. These interests motivate research question 2, "How do the characteristics of private high schools visited by public research universities compare to the characteristics of private high schools visited by private universities?"

Decisions about which private school characteristics to analyze were informed by prior research [e.g., @RN4504; @RN4321; @RN3519; @RN4588; @RN4724]. In particular, @RN4504 used data from several national surveys to analyze long-term trends in private school enrollment by family income. The percentage of American (elementary school) students attending private school declined from 15% in the mid-1960s to 10% in the mid-1970s, and declined gradually from 11% in 1999 to 9% in 2015. With respect to family income, the private school enrollment rate of high-income families (around 17%) and low-income families (around 4%) remained stable over time, but the enrollment rate for middle-income families declined from 12% in 1968 to 6% in 2013.  These declines were substantially driven by declines in the number of students attending Catholic school, which represented 89% of private school enrollment in 1965 and 42% of enrollment in 2013. Private school enrollment rates are higher for families located in urban areas than those in suburban areas. With respect to race, in 2013, 11% of white students attended private school compared to 5% of Black students, and 3% of Hispanic students. However, for high income families, rates of private school enrollment do not differ substantially across race. In the South, there has been a long-term increase in private school enrollment, driven by growing enrollment in Christian private schools by middle-income families and growing enrollment in nonsetarian private schools by high-income families.

Based on these findings and logical argument (and also data limitations), our analyses focus on the following private school characteristics: geographic region, religious affiliation, academic reputation, racial composition, and enrollment size. We analyze geographic region because, for most colleges and universities, enrollment demand is likely to be stronger at high schools in the same geographic region, but nationally prestigious colleges and universities may experience high demand throughout the country. We analyze religious affiliation because private colleges are more likely to have a relationship with private high schools with a similar religious affiliation. Although public universities are formally secular, we are interested in whether particular public universities tend to target religiously affiliated private high schools. We analyze academic reputation based on the idea that more highly ranked colleges are likely to target highly ranked private high schools. Racial composition is of substantive interest because the racial composition of visited high schools has consequences for the rational composition of colleges and universities. Given our interests in social reproduction versus social mobility, we would also analyze tuition price and household income, but these data are not avalable for private high schools.

@@

This chapter conceptualizes recruiting visits to private high schools as a behavior consistent with the social reproduction orientation of selective private colleges and universities. We are interested in how similar the recruiting networks of public research universities are to those of private colleges and universities, thereby motivating research question 3: to what extent do public research universities and private universities visit overlapping sets of private high schools?

@@

# Data and Methods

## Data

Our broader project collected data on off-campus recruiting visits made by a convenience sample of colleges and universities during the 2017 calendar year. The data collection sample comes from three different lists of postsecondary institutions: all public research-extensive universities as defined by the 2000 Carnegie Classification (N=102); all private universities in the top 100 of U.S. News and World Report National Universities rankings (N=58); and all private colleges in the top 50 of U.S. News and World Report Liberal Arts Colleges rankings (N=47). For each of these institutions, we investigated their admissions website for pages that posted their upcoming off-campus recruiting visits. For institutions that posted such pages, we scraped the pages once per week throughout 2017. Many colleges and university only posted certain kinds of events (e.g., hotel receptions and national college fairs) but not others (e.g., day-time visits to high schools). These institutions are excluded from the analyses. Our final analysis sample consists of 15 public research universities, 14 private research universities, and 12 private liberal arts colleges. <!--[MAYBE ADD TABLE/FIGURE OR APPENDIX TABLE/FIGURE ON CHARACTERISTICS OF INSTITUTIONS IN OUR SAMPLE?]-->

@@

Table \@ref(tab:sample-characteristics) shows selected characteristics for universities in our analysis sample. Given that we collected recruiting data for private universities ranked in the top 100 of US News and World Report, one concern is whether it makes sense to compare public research universities not ranked in the top 100 to private universities ranked in the top 100. However, the columns in Table \@ref(tab:sample-characteristics) for 25th and 75th percentile SAT score of enrolled freshmen suggest that private universities ranked lower than 75 have test score ranges that are fairly similar to public research universities not ranked in the top 100. The five public research universities ranked 124 or below have score ranges that are modestly lower, particularly the University of Nebraska-Lincoln, but remain comparable to private universities ranked 75 to 100. One exception is that Stevens Institute of Technology (rank 80) has substantially higher test scores.

CRYSTAL - ADD ONE BRIEF PARAGRAPH LISTING THE DIFFERENT SECONDARY DATA SOURCES WE USED

## Methods

@@

## Research question 1: scale of visits

Research question 1 asks, how does the scale of visits to private high schools by public research universities compare to the scale of visits to private high schools by selective private universities? This question can be answered by counting the number of visits to private high schools, which equivalent to calculating degree centrality for each university. In addition to scale, we are also interested in the relative scale of visits to private high schools. That is, the number of visits to private high schools as a proportion of all off-campus recruiting visits, or as a proportion of all recruiting visits to high schools.

>>>>

---
title: "Recruiting Chapter"
author: 
  - Ozan Jaquette
  - Crystal Han
  - Irma Castañeda
bibliography: ./../other/burd_recruiting_chapter.bib
output: 
  bookdown::pdf_document2:
    toc: FALSE
csl: ./../other/apa.csl
urlcolor: blue
header-includes:
      - \usepackage{pdflscape}
      - \usepackage{geometry}
      - \usepackage{floatrow}
      - \usepackage{setspace}
      #- \doublespacing
      - \floatsetup{capposition=top}
fontsize: 11pt
---

@@

Prior to the pandemic, admissions counselors from colleges and universities traveled to local high schools each year in order to meet prospective students and maintain relationships with guidance counselors at feeder schools [@RN3519]. @RN4721 analyzed off-campus recruiting visits made in 2017 by 15 public research universities. Surprisingly, 12 of the 15 universities made more visits to *out-of-state* high schools than *in-state* high schools. These out-of-state visits focused on public schools in affluent, predominantly white communities. Additionally, most public research universities in our sample made a disproportionate number of visits to out-of-state private high schools. For the University of California-Berkeley, for example, 155 (43%) of the 360 out-of-state high school visits were to private high schools. For the University of Colorado-Boulder, 321 (36%) of the 901 out-of-state high school visits were to private high schools.

<!--
Note: The data for this book chapter will be slightly different due to using 2017-18 CCD data plus modified criteria (no longer excluding HS w/ certain school types) - but % is still similar

UC Berkeley: 171 (42%) of 404 out-of-state HS visits are to private schools
CU Boulder: 381 (38%) of 1008 out-of-state HS visits are to private schools
-->

@@

The mission of selective private colleges and universities is more squarely associated with social reproduction rather than social mobility. Historically, these institutions depend on patronage from philanthropic interests -- in connection with business and religious interests -- and on wealthy households who provide tuition revenue and donations [@RN4517]. In turn, selective private institutions contribute to social reproduction by providing access to their patrons. The "Chetty" data on parental income by college show that the overwhelming majority of students at selective private institutions have parental income in the top quintile [@RN4722; @RN4723]. A significant caveat is that religiously affiliated universities may prioritize enrollment opportunities for students of the same faith regardless of wealth, but even here the mission is to benefit families from a particular cultural status group. <!-- [CITE] -->

The traditional, albeit mythologized, public good mission of public research universities is tied to social mobility and development of the state [@RN2269; @RN921]. Public research universities are designated the unique responsibility of educating the future civic, professional, and business leaders of the state. Quoting 19th century University of Michigan President James Angell, these institutions contribute to social mobility by providing "an uncommon education for the common man" [as cited in @RN3608, p. 279] who could not afford tuition at elite private institutions.

We argue that the recruiting behavior of colleges and universities is an indicator of organizational enrollment priorities and, in turn, the extent to which enrollment priorities serve social mobility or social reproduction. Recruiting visits to private high schools are consistent with the social reproduction mission of selective private universities, regardless of whether the university is targeting a particular high school because of its academic reputation, religious affiliation, or the affluence of its student body. By contrast, for public research universities, visits to private high schools -- particularly out-of-state schools -- are antithetical to the mission of social mobility for high-achieving, low-income state residents.

Analyses of recruiting visits to private high schools can contribute novel insights about the "privatization" of public research universities. Within the scholarly literature on privatization, one strand of scholarship conceptualizes privatization as a process whereby declines in government funding cause public institutions to grow alternative revenue sources, particularly revenue from tuition and research commercialization [@RN4543; @RN2101]. Of relevance to this chapter, public research universities responded to declines in state appropriations by growing enrollment from nonresident students [@RN3753], who typically pay two- to three-times more than resident students. A second strand within the privatization literature conceptualizes privatization as a process whereby public institutions become more similar -- in terms of characteristics or behavior -- to for-profit or private non-profit organizations [@RN4543]. A weakness of this literature is that few empirical studies analyze public institutions on a behavior associated with private institutions.

Therefore, this chapter compares public research universities to selective private universities on the dimension of recruiting visits to private high schools, a behavior associated with the social reproduction orientation of selective private universities. Our primary goal is to learn about public research universities, using the behavior of private universities as a benchmark to develop more thoughtful insights about public research universities.  We analyze recruiting visits made in the 2017 calendar year by a sample of 15 public research universities and 14 selective private universities. <!-- and 12 selective private liberal arts colleges. -->

Because little is known about recruiting at private high schools by public universities, we conduct descriptive analyses that seek to identify basic patterns and relationships. Because a recruiting visit to the same private high school is an indicator of a relationship between two actors, we also utilize social network methods in order to compare the recruiting networks of public universities to those of private universities. Analyses are informed by the following three research questions:

@@

<!-- SHORT PARAGRAPH GIVING OUTLINE OF CHAPTER -->

@@

## Situating off-campus recruiting within enrollment management

@@

Institutions identify undergraduate "prospects" by purchasing "student lists" -- containing contact, demographic, and academic achievement information -- from College Board, ACT, and other vendors. Universities control which prospects are contained within a list by filtering on criteria such as zip code, test scores, and high school GPA. Once prospects and inquiries are identified, they are targeted with remote and face-to-face recruiting interventions designed to solicit applications and deepen engagement. Remote recruiting interventions include postcards, brochures, emails, text messages, and targeted social media. Face-to-face recruiting interventions include on-campus visits by prospects and off-campus recruiting visits by admissions representatives to high schools, community colleges, college fairs, etc. Institutions utilize advertising (traditional and digital) and social media to solicit inquiries and to create positive "buzz" amongst prospects [@RN4411]<!--[CITE OTHER]-->. Given the rise in "stealth applicants" who do not inquire before applying (Dupaul & Harris, 2012), advertising and social media enables universities to tell their story to prospects who do not wish to be contacted.

@@

## Empirical scholarship from sociology

@@

Empirical academic scholarship on off-campus recruiting is mostly limited to a modest number of sociological case studies, which typically analyze off-campus recruiting as part of a broader analysis of enrollment management [e.g., @RN4538; @RN4324; @RN3519; @RN4520; @RN4407]. @RN4324 analyzes visits from the perspective of students at two racially and socioeconomically diverse public high schools. @RN4324 found that high school visits -- including college fairs, instant decision events, and small-group representative visits -- influenced where students applied and where they enrolled. This finding was strongest for first-generation students and underrepresented students of color, who often reported that "school counselors had low expectations for them and were too quick to suggest that they attend community college" (p. XXX). This trust vacuum created an opportunity for colleges because these students were drawn to colleges that connected with them and made them feel wanted. For example, @RN4324 describes a high-achieving, first-generation, African American student who was admitted by a highly selective liberal arts college but chose to attend a less selective college that "seemed to want him more" (p. XXX). By contrast, affluent students with college-educated parents were less taken by such overtures and more concerned with college prestige.

@RN3519, in an ethnography of the admissions office at a selective private liberal arts college, highlights the relational function of off-campus recruiting visits. Stevens [-@RN3519, p. 54] states that "the College's reputation and the quality of its applicant pool are dependent upon its connections with high schools nationwide." Therefore, during the autumn "travel season," admissions officers visit selected high schools across the country "to spread word of the institution and maintain relationships with guidance counselors" (p. 53-54). The College tended to visit the same "feeder" schools year after year because recruiting depends on long-term relationships with high schools. The high schools they visited tend to be affluent schools -- in particular, private schools -- that enroll high-achieving students who can afford tuition and have the resources and motivation to host a successful visit. Whereas @RN4402 highlights the effect of recruiting visits on inquiries and enrollees, findings from @RN3519 suggest that the College may have valued recruiting visits primarily as a means of maintaining relationships with guidance counselors. From this perspective, recruiting visits may affect outcomes such as inquiries, applications, and matriculation through their effect on high school guidance counselors. The logic is that a guidance counselor who views a college favorably will steer students to the college.

@@

The horsetrading described by @RN4407 depends on a relationship where the college can trust statements made by the high school counselor and vice-versa. This relationship is the product of repeated interactions over many years. A high school that makes false statements about applicant intentions faces consequences. The college "might stop taking these telephone calls and ignore the information provided. They may even start accepting fewer students from the school, thinking it is less than an honest place" [@RN4321, p. 106]. Because the college and the high school are mutually dependent, however, both "have an incentive to continue with a strong, honest relationship" [@RN4321, p. 106]. Although, such horsetrading may be less pervasive now than it was in prior decades and may be less common at non-elite private schools. Additionally, college admissions officers from public universities likely have less authority to engage in these tacit negotiations because admissions criteria at public universities face greater public scrutiny than admissions criteria at private colleges and universities.

@@

@RN3519 and Khan [-@RN4321;-@RN4407] suggest that strong relationships are mutually important for the college and for the private high school. Strong relationships enable colleges and schools to negotiate and send trustworthy information to one another. Without a strong relationship, it is unlikely that a college admissions counselor will "take the call" of a high school guidance counselor [@RN4321; @RN4407]. Even in the absence of the horsetrading described by @RN4407, relationships enable the college admissions counselor and the high school guidance counselor to tell one another "their story" and relay that story to constituents [@RN3519]. The college admissions counselor explains why the college/university is a "special place" that high school students should want to attend [@RN3519]. The guidance counselor explains why the high school is a special place, that even students with lower grades have outstanding extracurricular strengths that will benefit the college [@RN4321].

@@

Having conceptualized a recruiting visit as an indicator of a social relation, this section introduces basic concepts from social network analysis and motivates the research questions that guide empirical analyses.

@@

Whereas "one-mode" networks consist of vertices of the same "type" (e.g., in a publication network, each vertex is an author), "two-mode" networks consist of vertices associated with one type of actor/entity having connections to vertices of another type. For example, an actor-movie network consists of actors (mode 1) who appear in movies (mode 2), and an actor shares an edge with a movie if the actor appears in the movie. The most commonly analyzed two-mode network in social networks literature is a corporate board-director network  [e.g., @RN4663]. These networks consist of directors (mode 1) and organizational boards (mode 2), and a particular director shares an edge with a particular organizational board if they sit on that board. Similarly, our recruiting visit data form a two-mode network, consisting of private high schools (mode 1) and colleges (mode 2). An off-campus recruiting visit from a college/university to a private high school is the "edge" that connects two particular vertices. In this network, edges can only occur between vertices of a different type; that is, a college can visit a high school, but colleges cannot visit other colleges and high schools cannot receive visits from other high schools.

@@

In social network analysis, *centrality* refers to the importance or prominence or "being in the middle of things" of an actor/vertex in a social network.<!--[CITE]--> Many alternative measures of centrality have been developed. All measures agree that centrality is an actor/vertex-level construct (e.g., each college in our network will have a centrality value, regardless of which centrality measure we use). However, different measures of centrality are based on different theoretical conceptualizations -- for example, being directly connected to the most actors versus being the intermediary between otherwise disconnected actors -- and are operationalized using different mathematical algorithms.

The simplest measure of centrality, *degree centrality* measures the number of edges directly connected to a vertex. For each college in our two-mode network, degree centrality is simply the number of different private high schools visited by the college. For each high school in our network, degree centrality is the number of colleges in our network that visited the high school. Social network theory conceptualizes degree centrality as a measure of "local" centrality, capturing the number of direct connections for each actor/vertex. For our empirical context, we conceive of degree centrality as a measure of scale of the number of private high schools a college visits and, equivalently, the number of private high schools we observe having a social relation with each college. Given that we conceptualize visits to private high schools as a behavior consistent with the social reproduction orientation of selective private colleges/universities, we are interested in comparing how scale differs between public and private colleges/universities. This interest motivates research question 1, "How does the scale of visits to private high schools by public research universities compare to the scale of visits to private high schools by selective private universities?" We answer this research question by comparing degree centrality across universities.

@@

Degree is a limited measure of centrality in that it only identifies direct connections. By contrast, *k-path centrality* measures the extent to which an actor/vertex is indirectly connected to others, *betweenness centrality* measures the extent to which a vertex is an intermediary between otherwise disconnected groups of actors, and *eigenvector centrality* measures the extent to which a vertex is connected to highly-connected others. Betweenness centrality tends to be relevant for empirical contexts (e.g., which manager gets credit for an idea) where an actor who occupies a position between groups can benefit by being a broker of information/ideas [@RN975]. By contrast, this chapter is more concerned with whether a particular college tends to have indirect connections with some colleges but not others, and what is the subset of high schools this group of colleges tends to visit. Research question 3, motivated below, explores these ideas.

@@

A common finding in social networks research is that actors are more likely to form connections with actors who are similar to them [@RN1722]. Said differently, ego networks are biased towards "homophily" rather than "heterophily." Homophily is the idea that two vertices are more likely to be connected if they have similar characteristics (e.g., two people are more likely to be friends on Facebook if they have similar political ideologies). Heterophily, the opposite of homophily, is when pairs of connected vertices have different characteristics from one another.

In our analyses of recruiting visits from colleges to private high schools, we are interested in the existence of homophily versus heterophily for two reasons. First, prior scholarship suggests that private colleges and private high schools are more likely to have a relationship when they share common ideologies or view themselves as belonging to the same echelon [@RN3519; @RN4321;@RN4407]. Second, with respect to our substantive interest in the behavior of public research universities, the characteristics of visited private high schools have consequences for subsequent enrollment compositions of public research universities. Therefore, we are interested in whether universities tend to visit high schools that are similar to them on certain characteristics. Additionally, we are interested in whether public research universities favor high schools with certain characteristics, independent of homophily. These interests motivate research question 2, "How do the characteristics of private high schools visited by public research universities compare to the characteristics of private high schools visited by private universities?"

Decisions about which private school characteristics to analyze were informed by prior research [e.g., @RN4724; @RN4504; @RN4321; @RN3519; @RN4588]. In particular, @RN4504 used data from several national surveys to analyze long-term trends in private school enrollment by family income. The percentage of American (elementary school) students attending private school declined from 15% in the mid-1960s to 10% in the mid-1970s, and declined gradually from 11% in 1999 to 9% in 2015. With respect to family income, the private school enrollment rate of high-income families (around 17%) and low-income families (around 4%) remained stable over time, but the enrollment rate for middle-income families declined from 12% in 1968 to 6% in 2013.  These declines were substantially driven by declines in the number of students attending Catholic schools, which represented 89% of private school enrollment in 1965 and 42% of enrollment in 2013. Private school enrollment rates are higher for families located in urban areas than those in suburban areas. With respect to race, in 2013, 11% of white students attended private school compared to 5% of Black students, and 3% of Hispanic students. However, for high income families, rates of private school enrollment do not differ substantially across race. In the South, there has been a long-term increase in private school enrollment, driven by growing enrollment in Christian private schools by middle-income families and growing enrollment in nonsectarian private schools by high-income families.

Based on these findings and logical argument (and also data limitations), our analyses focus on the following private school characteristics: geographic region, religious affiliation, academic reputation, racial composition, and enrollment size. We analyze geographic region because, for most colleges and universities, enrollment demand is likely to be stronger at high schools in the same geographic region, but nationally prestigious colleges and universities may experience high demand throughout the country. We analyze religious affiliation because private colleges are more likely to have a relationship with private high schools with a similar religious affiliation. Although public universities are formally secular, we are interested in whether particular public universities tend to target religiously affiliated private high schools. We analyze academic reputation based on the idea that more highly ranked colleges are likely to target highly ranked private high schools. Racial composition is of substantive interest because the racial composition of visited high schools has consequences for the racial composition of colleges and universities. Given our interests in social reproduction versus social mobility, we would also analyze tuition price and household income, but these data are not available for private high schools.

@@

This chapter conceptualizes recruiting visits to private high schools as a behavior consistent with the social reproduction orientation of selective private colleges and universities. We are interested in how similar the recruiting networks of public research universities are to those of private colleges and universities, thereby motivating research question 3, "To what extent do public research universities and private universities visit overlapping sets of private high schools?"

@@

# Data and methods

Our broader project collected data on off-campus recruiting visits made by a convenience sample of colleges and universities during the 2017 calendar year. The data collection sample comes from two different lists of postsecondary institutions: all public research-extensive universities as defined by the 2000 Carnegie Classification (N=102), and all private universities in the top 100 of U.S. News and World Report National Universities rankings (N=58). <!-- and all private colleges in the top 50 of U.S. News and World Report Liberal Arts Colleges rankings (N=47).-->For each of these institutions, we investigated their admissions website for pages that posted their upcoming off-campus recruiting visits. For institutions that posted such pages, we scraped the pages once per week throughout 2017. Many colleges and universities only posted certain kinds of events (e.g., hotel receptions and national college fairs) but not others (e.g., day-time visits to high schools). These institutions are excluded from the analyses. Our final analysis sample consists of 15 public research universities and 14 private research universities. <!--12 private liberal arts colleges [MAYBE ADD TABLE/FIGURE OR APPENDIX TABLE/FIGURE ON CHARACTERISTICS OF INSTITUTIONS IN OUR SAMPLE?]-->

@@

Table \@ref(tab:sample-characteristics) shows selected characteristics for the universities in our analysis sample. Given that we collected recruiting data for private universities ranked in the top 100 of U.S. News and World Report, one concern is whether it makes sense to compare public research universities not ranked in the top 100 to private universities ranked in the top 100. However, the columns in Table \@ref(tab:sample-characteristics) for 25th and 75th percentile SAT score of enrolled freshmen suggest that private universities ranked lower than 75 have test score ranges that are fairly similar to public research universities not ranked in the top 100. The five public research universities ranked 124 or below have score ranges that are modestly lower, particularly the University of Nebraska-Lincoln, but remain comparable to private universities ranked 75 to 100. One exception is that Stevens Institute of Technology (rank 80) has substantially higher test scores.

The data on university characteristics come from the 2017 Integrated Postsecondary Education Data System (IPEDS) and 2020 Best Colleges ranking by U.S. News & World Report. For the high schools in our analysis, we use secondary data from the 2017-18 NCES Common Core of Data (CCD) for public high schools, and the 2017-18 NCES Private School Universe Survey (PSS) and 2020 Best Private High Schools ranking by Niche for private high schools.^[For both public and private high schools, we also used NCES data from previous years when the 2017-18 data was not available for a small amount of schools.] Only high schools that enroll at least 10 12th grade students, is not primarily a virtual school, and is located in the 50 U.S. states, the District of Columbia, or land regulated by the Bureau of Indian Affairs is included in our analysis.

@@

## RQ1: Scale of visits

Research question 1 asks, "How does the scale of visits to private high schools by public research universities compare to the scale of visits to private high schools by selective private universities?" This question can be answered by counting the number of visits to private high schools, which is equivalent to calculating the degree centrality for each university. In addition to scale, we are also interested in the relative scale of visits to private high schools. That is, the number of visits to private high schools as a proportion of all off-campus recruiting visits, or as a proportion of all recruiting visits to high schools.
